---
title: "inference.hw"
author: Karli Balzanto
format: html
embed-resources: true
---

## problem 1
explicit hypotheses
- Null Hypothesis: $\mu = 110$
- Alternative: $\mu \ne 110$ ($H_a: \mu \ne 110$)

conclusion
- the p-value of 0.000001429 means that data like ours (a sample mean this far from 110) would only occur about 0.00014% of the time if the true average price really were $110.
- The null hypothesis doesn’t seem compatible with our data. We have strong evidence against the null hypothesis, so we reject it and conclude that the alternative is true. 
```{r}
#look at info
summary(ISLR2::Carseats$Price)
t.test(ISLR2::Carseats$Price,
       mu = 110)

#visual
library(ISLR2)
library(ggplot2)

ggplot(Carseats,
       aes(x = Price)) +
  geom_histogram(binwidth = 5, 
                 fill = "skyblue", 
                 color = "white", 
                 alpha = 0.8) +
  geom_vline(xintercept = 110, 
             color = "pink", 
             size = 1.2) +
  geom_vline(xintercept = mean(Carseats$Price), 
             color = "lightgreen", 
             size = 1.2) +
  labs(title = "Distribution of Car Seat Prices",
       subtitle = "Red = Claimed | Green = Sample",
       x = "Price ($)",
       y = "Count") +
  theme_minimal()

```


## problem 2
```{r}
xbar <- mean(Carseats$CompPrice)
s <- sd(Carseats$CompPrice)
n <- nrow(Carseats)
tstar <- qt(0.95, df = n-1)

margin <- tstar * s / sqrt(n)
ci <- c(xbar - margin, xbar + margin)

xbar      # sample mean: 124.975
margin    # margin error: 1.264086
ci        # 90%: 123.7109  126.2391

```


## problem 3
```{r}
penguins <- addmargins(table(penguins$species,
                             penguins$island))
penguins
```


## problem 4

couldn't push before due date so went to office hours and fixed. added [link](https://github.com/karb-05/MATH250_github_sandbox) after.

