---
title: "Karli_Balzanto_Final"
author: "Karli Balzanto"
format: html
embed-resources: true
---
```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(ggplot2)
library(dplyr)
library(gt)
college <- read.csv("college.csv")
pressure <- read.csv("pressure.csv")
billboard <- read.csv("billboard.csv")
```

# Final Exam
Followed all best practices from the entire semester, showcasing skills gained through this class.

### problem 1

*hypotheses*

- $H_0:$ Mean systolic blood pressure is the same across lifestyle groups (urban, rural, suburban).
- $H_a:$ At least one lifestyle group has a different mean systolic pressure.

```{r}
#| label: question 1
#| fig-alt: xx
#| warning: false

model1 <- aov(systolic ~ lifestyle, data = pressure)
summary(model1)
```

*conclusion*

- The ANOVA test produced a very small p-value (p < 0.0001). At $\alpha = 0.5$ we reject $H_0$.
- Therefore, there is strong statistical evidence that systolic blood pressure is associated with lifestyle in this Ohio sample.




### problem 2

*hypotheses*

- $H_0: \mu = 128$ (Ohio mean systolic equals U.S. mean)
- $H_a: \mu \ne 128$

```{r}
#| label: question 2
#| fig-alt: xx
#| warning: false

t.test(pressure$systolic, mu = 128)
```

*conclusion*

- The test produced a very small p-value (p < 0.001). At $\alpha = 0.5$, we reject $H_0$.
- Therefore, the data provides strong evidence that Ohioans differ from the national average systolic blood pressure of 128.




### problem 3

*hypotheses*

- $H_0: \beta_1 = 0$ (acceptance rate does not explain graduation rate)
- $H_1: \beta_1 \ne 0$

```{r}
#| label: question 3
#| fig-alt: xx
#| warning: false

college$Accept <- as.character(college$Accept)

for (i in 1:length(college$Accept)) {
  value <- college$Accept[i]
  percent_number <- substring(value, 1, nchar(value) - 1)
  college$Accept[i] <- as.numeric(percent_number) / 100}

college$Accept <- as.numeric(college$Accept)
college$Grad.Rate <- as.numeric(college$Grad.Rate)
college$Grad.Rate <- as.numeric(college$Grad.Rate)

model2 <- lm(Grad.Rate ~ Accept, data = college)
summary(model2)
```

```{r}
#| label: question 3 pt 2
#| fig-alt: xx
#| warning: false

ggplot(college,
       aes(x = Accept,
           y = Grad.Rate)) +
  geom_point(color = "lightblue") +
  geom_smooth(method = "lm", color = "black") +
  labs(title = "Graduation Rate vs Acceptance Rate",
       x = "Acceptance Rate (fraction)",
       y = "Graduation Rate") +
  theme_minimal()
```


*conclusion*

- The regression output shows that the slop is statistically significant with a very small p-value. At $\alpha = 0.5$, we reject $H_0$.
- Therefore, there's strong evidence that acceptance rate is a meaningful predictor of graduation rate.




### problem 4

*equation*

- $GradRate = \beta_0 + [\beta_1 * Accept] + [\beta_2 * Public]$
  - where 1 for public & 0 for private

```{r}
#| label: question 4
#| fig-alt: xx
#| warning: false

college$Public <- NA
college$Public[college$Private == "Yes"] <- 0 #private
college$Public[college$Private == "No"]  <- 1 #public

model3 <- lm(Grad.Rate ~ Accept + Public, data = college)
summary(model3)
```


*interpretation*

- $\beta_0$: Predicted graduation rate for a private college with 0 acceptance rate.
- $\beta_1$: Change in graduation rate for every unit increase in acceptance rate.
- $\beta_2$: Difference between public and private colleges, holding acceptance rate constant.(public - private)

```{r}
#| label: question 4 pt 2
#| fig-alt: xx
#| warning: false

ggplot(college,
       aes(x = Accept,
           y = Grad.Rate,
           color = factor(Public))) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_brewer(palette = "Dark2") +
  labs(color = "Public (1) / Private (0)",
       title = "Graduation Rate") +
  theme_minimal()
```

*predicted graduation rate*


```{r}
#| label: question 4 pt 3
#| fig-alt: xx
#| warning: false

b0 <- coef(model3)[1]
b1 <- coef(model3)[2]
b2 <- coef(model3)[3]

#accept val = 0.85
#public val = 0
predicted_value <- b0 + b1 * 0.85 + b2 * 0
predicted_value
```


### problem 5

*write-up*

- I dug into the Billboard #1 songs data set to see how songs reaching the top varied across years and decades. The data shows that most #1 hits only stay at the top for a few weeks, but a few songs break out and stay way longer — those tracks stand out in almost any decade. When I grouped the songs by decade, I noticed some decades have more entries than others; that likely comes from how the charts were tracked and how music was released over time. I picked the song with the longest run at #1 in the data set and highlighted it on the scatter so you can see how unusual it is compared to most #1's. Overall, the data set tells a story of consistent short-term hits with rare long-term blockbusters.

*visuals*
```{r}
#| label: question 5 visual 1
#| fig-alt: xx
#| warning: false

#1: Number of #1 Songs per Decade
billboard$date <- as.character(billboard$date)
billboard$year <- rep(NA, length(billboard$date))
for (i in 1:length(billboard$date)) {
  billboard$year[i] <- as.numeric(substring(billboard$date[i], 1, 4))}

billboard$decade <- (billboard$year %/% 10) * 10

ggplot(billboard,
       aes(x = decade,
           fill = factor(decade))) +
  geom_bar(show.legend = FALSE) +
  labs(title = "Number of Billboard #1 Songs per Decade",
       x = "Decade",
       y = "Count") +
  scale_fill_brewer(palette = "Dark2") +
  theme_minimal()
```


```{r}
#| label: question 5 visual 2
#| fig-alt: xx
#| warning: false

#2: Weeks at #1 Over Time with Highlighted Song
highlight <- billboard[order(-billboard$weeks_at_number_one), ][1, ]

ggplot(billboard,
       aes(x = year,
           y = weeks_at_number_one)) +
  geom_point(alpha = 0.5) +
  geom_point(data = highlight,
             color = "red",
             size = 3) +
  annotate("text",
           x = highlight$year + 1,
           y = highlight$weeks_at_number_one + 1,
           label = highlight$song,
           color = "red") +
  labs(title = "Weeks at #1 Over Time",
       x = "Year",
       y = "Weeks at #1") +
  theme_minimal()
```

```{r}
#| label: question 5 visual 3
#| fig-alt: xx
#| warning: false

#3: Top 10 Longest #1 Songs
top10 <- billboard %>%
  arrange(desc(weeks_at_number_one)) %>%
  select(song, artist, weeks_at_number_one) %>%
  slice(1:10) %>%
  gt()
top10
```

### problem 6

*GitHub*
put problem 5 into GitHub


